---
title: "HW02p"
author: "[Your Name Goes Here]"
date: "March 6, 2018"
output: pdf_document
---

```{r setup, cache = F}
knitr::opts_chunk$set(error = TRUE) #this allows errors to be printed into the PDF
```

Welcome to HW02p where the "p" stands for "practice" meaning you will use R to solve practical problems. This homework is due 11:59 PM Tuesday 3/6/18. 

You should have RStudio installed to edit this file. You will write code in places marked "TO-DO" to complete the problems. Some of this will be a pure programming assignment. Sometimes you will have to also write English.

The tools for the solutions to these problems can be found in the class practice lectures. I want you to use the methods I taught you, not for you to google and come up with whatever works. You won't learn that way.

To "hand in" the homework, you should compile or publish this file into a PDF that includes output of your code. To do so, use the knit menu in RStudio. You will need LaTeX installed on your computer. See the email announcement I sent out about this. Once it's done, push the PDF file to your github class repository by the deadline. You can choose to make this respository private.

For this homework, you will need the `testthat` libray.

```{r}
pacman::p_load(testthat)
```


1. Source the simple dataset from lecture 6p:

```{r}
Xy_simple = data.frame(
 response = factor(c(0, 0, 0, 1, 1, 1)), #nominal
 first_feature = c(1, 1, 2, 3, 3, 4),    #continuous
 second_feature = c(1, 2, 1, 3, 4, 3)    #continuous
)
X_simple_feature_matrix = as.matrix(Xy_simple[, 2 : 3])
y_binary = as.numeric(Xy_simple$response == 1)


```


Try your best to write a general perceptron learning algorithm to the following `Roxygen` spec. For inspiration, see the one I wrote in lecture 6.

```{r}
#' This function implements the "perceptron learning algorithm" of Frank Rosenblatt (1957).
#'
#' @param Xinput      The training data features as an n x (p + 1) matrix where the first column is all 1's.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param MAX_ITER    The maximum number of iterations the perceptron algorithm performs. Defaults to 1000.
#' @param w           A vector of length p + 1 specifying the parameter (weight) starting point. Default is 
#'                    \code{NULL} which means the function employs random standard uniform values.
#' @return            The computed final parameter (weight) as a vector of length p + 1
perceptron_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 1000, w = NULL){
  if(is.null(w)) { #if w is empty
    w = runif(ncol(Xinput)) #fill w with n+1 random numbers
  }
  for(iter in 1:MAX_ITER){#perfom alogirithm 1000 times
    for(i in 1:nrow(Xinput)){#loopd over each piece of data in features matrix
      x_i = Xinput[i, ] #set x_i equal to the ith piece of data with all its' columns 
      yhat_i = ifelse(x_i %*% w > 0, 1, 0)
      w = w + as.numeric(y_binary[i] - yhat_i) * x_i #update w
    }
  } 
w
}
```

Run the code on the simple dataset above via:

```{r}
w_vec_simple_per = perceptron_learning_algorithm(
  cbind(1, Xy_simple$first_feature, Xy_simple$second_feature),
  as.numeric(Xy_simple$response == 1))
w_vec_simple_per
```

Use the ggplot code to plot the data and the perceptron's $g$ function.

```{r}
pacman::p_load(ggplot2)
simple_viz_obj = ggplot(Xy_simple, aes(x = first_feature, y = second_feature, color = response)) + geom_point(size = 5)
simple_perceptron_line = geom_abline(
    intercept = -w_vec_simple_per[1] / w_vec_simple_per[3], 
    slope = -w_vec_simple_per[2] / w_vec_simple_per[3], 
    color = "orange", show.legend = TRUE)
simple_viz_obj + simple_perceptron_line
```

Why is this line of separation not "satisfying" to you?
We can easily draw a better line
Everytime we run the PLA it will give us a different line so the PLA has no concept of what is the "best" line.

TO-DO



2. Use the `e1071` package to fit an SVM model to `y_binary` using the predictors found in `X_simple_feature_matrix`. Do not specify the $\lambda$ (i.e. do not specify the `cost` argument).

```{r}
pacman::p_load(e1071)

svm_model = svm(X_simple_feature_matrix, y_binary, kernel = "linear", scale = FALSE)

```

and then use the following code to visualize the line in purple:

```{r}
w_vec_simple_svm = c(
  svm_model$rho, #the b term
  -t(svm_model$coefs) %*% X_simple_feature_matrix[svm_model$index, ] # the other terms
)
w_vec_simple_svm
simple_svm_line = geom_abline(
    intercept = -w_vec_simple_svm[1] / w_vec_simple_svm[3], 
    slope = -w_vec_simple_svm[2] / w_vec_simple_svm[3], 
    color = "purple")
simple_viz_obj + simple_perceptron_line + simple_svm_line

```

Is this SVM line a better fit than the perceptron?
It's worse since it doesn't properly seperate the data

3. Now write your own implementation of the linear support vector machine algorithm respecting the following spec making use of the nelder mead `optim` function from lecture 5p. It turns out you do not need to load the package `neldermead` to use this function. You can feel free to define a function within this function if you wish. 

Note there are differences between this spec and the perceptron learning algorithm spec in question \#1. You should figure out a way to respect the `MAX_ITER` argument value. 

```{r}
#' This function implements the hinge-loss + maximum margin linear support vector machine algorithm of Vladimir Vapnik (1963).
#'
#' @param Xinput      The training data features as an n x p matrix.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param MAX_ITER    The maximum number of iterations the algorithm performs. Defaults to 5000.
#' @param lambda      A scalar hyperparameter trading off margin of the hyperplane versus average hinge loss.
#'                    The default value is 0.1 to mimic hard margin in the linearly separable case.
#' @return            The computed final parameter (weight) as a vector of length p + 1
linear_svm_learning_algorithm = function(Xinput, y_binary, MAX_ITER = 5000, lambda = 0.1){
  #TO-DO
}
```

Run your function using the defaults and plot it in brown vis-a-vis the previous model's line:

```{r}
svm_model_weights = linear_svm_learning_algorithm(X_simple_feature_matrix, y_binary)
my_svm_line = geom_abline(
    intercept = -svm_model_weights[1] / svm_model_weights[3], 
    slope = -svm_model_weights[2] / svm_model_weights[3], 
    color = "brown")
simple_viz_obj + simple_svm_line + my_svm_line
```

Is this the same as what the `e1071` implementation returned? Why or why not?

4. Write a $k=1$ nearest neighbor algorithm using the Euclidean distance function. Respect the spec below:

```{r}
#' This function implements the nearest neighbor algorithm.
#'
#' @param Xinput      The training data features as an n x p matrix.
#' @param y_binary    The training data responses as a vector of length n consisting of only 0's and 1's.
#' @param Xtest       The test data that the algorithm will predict on as a n* x p matrix.
#' @return            The predictions as a n* length vector.
nn_algorithm_predict = function(Xinput, y_binary, Xtest){
  best_sqd = Inf #initiate the best squared distance to a huge value
  for(i in 1:nrow(Xinput)) { #loop over rows
    dsqd = 0 #reset distance squared to 0 each time we start looping on a new row
    for(j in 1:ncol(Xinput)) { #loop over columns
      dsqd = dsqd + (Xtest[i, j] - Xinput[i, j])^2 #add the squared distance between Xinput at position i,j and Xtest at i,j
    }
    euclidean_distance = sqrt(dsqd) #calculate the euclidean distance between i'th row of input and test
    if(euclidean_distance < best_sqd) { #if euclidean distance is less than the best_sqd we already have..
      best_sqd = euclidean_distance # set best_sqd equal to the euclidean_distance
      best_row_index = i #keep track of which row has smallest distance
    }
  }
  y_binary[best_row_index]
}
```

Write a few tests to ensure it actually works:
```{r}
Xy = na.omit(MASS::biopsy) #The "breast cancer" data with all observations with missing values dropped
X = Xy[, 2 : 10] #V1, V2, ..., V9
y_binary = as.numeric(Xy$class == "malignant")
```

```{r}
xx = nn_algorithm_predict(X, y_binary, X)
xx
```

For extra credit, add an argument `k` to the `nn_algorithm_predict` function and update the implementation so it performs KNN. In the case of a tie, choose $\hat{y}$ randomly. Set the default `k` to be the square root of the size of $\mathcal{D}$ which is an empirical rule-of-thumb popularized by the "Pattern Classification" book by Duda, Hart and Stork (2007). Also, alter the documentation in the appropriate places.

```{r}
#not required TO-DO --- only for extra credit
```

For extra credit, in addition to the argument `k`, add an argument `d` representing any legal distance function to the `nn_algorithm_predict` function. Update the implementation so it performs KNN using that distance function. Set the default function to be the Euclidean distance in the original function. Also, alter the documentation in the appropriate places.

```{r}
#not required TO-DO --- only for extra credit
```


5. We move on to simple linear modeling using the ordinary least squares algorithm.

Let's quickly recreate the sample data set from practice lecture 7:

```{r}
n = 20
x = runif(n)
beta_0 = 3
beta_1 = -2
y = beta_0 + beta_1 * x + rnorm(n, mean = 0, sd = 0.33)

```

Solve for the least squares line by computing $b_0$ and $b_1$ _without_ using the functions `cor`, `cov`, `var`, `sd` but instead computing it from the $x$ and $y$ quantities manually. See the class notes.

```{r}
ybar = (1/n) * sum(y)
xbar = 1/n * sum(x)

b_1 = (sum(x*y)-n*xbar*ybar)/(sum(x^2)-n*xbar^2)
b_0 = ybar - b_1*xbar

g = b_0 + b_1*x

```

Verify your computations are correct using the `lm` function in R:

```{r}
lm_mod = lm(y ~x)
b_vec = coef(lm_mod)
expect_equal(b_0, as.numeric(b_vec[1]), tol = 1e-4) #thanks to Rachel for spotting this bug - the b_vec must be casted to numeric
expect_equal(b_1, as.numeric(b_vec[2]), tol = 1e-4)
```

6. We are now going to repeat one of the first linear model building exercises in history --- that of Sir Francis Galton in 1886. First load up package `HistData`.

```{r}
pacman::p_load(HistData)
```

In it, there is a dataset called `Galton`. Load it using the `data` command:

```{r}
data("Galton")

```

You now should have a data frame in your workspace called `Galton`. Summarize this data frame and write a few sentences about what you see. Make sure you report $n$, $p$ and a bit about what the columns represent and how the data was measured. See the help file `?Galton`.

n=928, p=2
This data set is made up of 928 observations of parent height and their childs height. Parent height is the average of the mother & father height. Female childs height was multipled by 1.08 before it was recorded to account for a difference in sex. Glaton used classes to group the data into, where each class interal has a width of 1.0 inch.


```{r}
summary(Galton)
```

TO-DO

Find the average height (include both parents and children in this computation).

```{r}


avg_height = (mean(Galton$parent) + mean(Galton$child))/2

```

Note that in Math 241 you learned that the sample average is an estimate of the "mean", the population expected value of height. We will call the average the "mean" going forward since it is probably correct to the nearest tenth of an inch with this amount of data.

Run a linear model attempting to explain the childrens' height using the parents' height. Use `lm` and use the R formula notation. Compute and report $b_0$, $b_1$, RMSE and $R^2$. Use the correct units to report these quantities.

```{r}
gal = lm(Galton$child ~ Galton$parent)
coefficients(gal)#returns b0 and b1
summary(gal)$r.squared#rsquraed
summary(gal)$sigma

```

Interpret all four quantities: $b_0$, $b_1$, RMSE and $R^2$.
b_0=23.94
b1=.65 for every addition inch in parents height, we can expect the childrens height to increase by .65 inches
R^2=.21 -the independent variables don't explain the dependent variables well
RMSE=2.24-our model is not very accurate when predicting childrens height


How good is this model? How well does it predict? Discuss.
This is a poor model since the R^2 is low and the RMSE is high. An RMSE value of 2.24 means our predictions are way off from the actual data.

TO-DO

Now use the code from practice lecture 8 to plot the data and a best fit line using package `ggplot2`. Don't forget to load the library.

```{r}
pacman::p_load(ggplot2)
ggplot(data = Galton, aes(x = parent, child)) + geom_point() + geom_abline(intercept = 23.94, slope = .65)
```

It is reasonable to assume that parents and their children have the same height. Explain why this is reasonable using basic biology.
it would make logical sense that the offspring of 2 people would have all the same features averaged together.
TO-DO

If they were to have the same height and any differences were just random noise with expectation 0, what would the values of $\beta_0$ and $\beta_1$ be?
beta=0, b1=x
TO-DO

Let's plot (a) the data in $\mathbb{D}$ as black dots, (b) your least squares line defined by $b_0$ and $b_1$ in blue, (c) the theoretical line $\beta_0$ and $\beta_1$ if the parent-child height equality held in red and (d) the mean height in green.

```{r}
ggplot(Galton, aes(x = parent, y = child)) + 
  geom_point() + 
  geom_jitter() +
  geom_abline(intercept = 23.94, slope = .65, color = "blue", size = 1) +
  geom_abline(intercept = 0, slope = 1, color = "red", size = 1) +
  geom_abline(intercept = avg_height, slope = 0, color = "darkgreen", size = 1) +
  xlim(63.5, 72.5) + 
  ylim(63.5, 72.5) +
  coord_equal(ratio = 1)
```

Fill in the following sentence: 

TO-DO: Children of short parents became short on average and children of tall parents became tall on average.

Why did Galton call it "Regression towards mediocrity in hereditary stature" which was later shortened to "regression to the mean"?


TO-DO

Why should this effect be real?

TO-DO

You now have unlocked the mystery. Why is it that when modeling with $y$ continuous, everyone calls it "regression"? Write a better, more descriptive and appropriate name for building predictive models with $y$ continuous.

TO-DO

